---
title: "Reproducible Data Peer Assignment 1"
author: "Richard Ryder"
date: "8/3/2020"
output: html_document
---

###  Assuming 'activity.csv' is in working directory load and process the data.
```{r}
     activity <- read.csv("activity.csv")
```
     
###  Create initial dataframes
 - dailysteps - total amount of steps per day
 - intervalsteps - average amount of steps per time interval
```{r}
     dailysteps <- aggregate(list(steps=activity$steps), 
          by = list(date = activity$date), 
          FUN = "sum", 
          na.rm = TRUE)
     
     intervalsteps <- aggregate(list(steps=activity$steps), 
          by = list(interval = activity$interval), 
          FUN = "mean", 
          na.rm = TRUE)
```
 
## What are the mean and median number of steps taken per day?
### Mean number of steps per day
```{r}
     mean <- mean(dailysteps$steps, na.rm=TRUE)
     mean
```
### Median number of steps per day
```{r}
     median <- median(dailysteps$steps, na.rm=TRUE)
     median
```
### Histogram of number of steps per day
```{r}
     hist(dailysteps$steps, 
          xlab = "# Daily Steps", 
          main="Frequency of Daily Steps")
```

## What is the average daily activity pattern?
```{r}
     plot(intervalsteps$interval, intervalsteps$steps, 
          type="l", 
          xlab="Minute intervals", 
          ylab="Steps (mean)")
```

### Which 5-minute interval, on average, contains the maximum number of steps?
```{r}
     intervalsteps[which.max(intervalsteps[,2]),]
```

## Inputing missing values

### How many NAs are in the original dataset?
```{r}
     sapply(activity, function(x) sum(is.na(x)))
```

### Create a copy of the original data set and replace NAs with the average interval steps previously calculated
 
```{r}
     newactivity <- activity
     newactivity[is.na(newactivity)] = intervalsteps$step
```

### Create an aggregate of the new dataset, then calculate new mean, median and histogram.
```{r}
     newsteps <- aggregate(list(steps=newactivity$steps), 
          by = list(date = newactivity$date), 
          FUN = "sum", 
          na.rm=TRUE)
```

### Original and new mean number of steps per day.
```{r}
     mean
     newmean <- mean(newsteps$steps, na.rm=TRUE)
     newmean
```

### Original and new median number of steps per day.
```{r}
     median
     newmedian <- median(newsteps$steps, na.rm=TRUE)
     newmedian
```

### Using updated data generate a histogram of the number of steps per day.
```{r}
     hist(newsteps$steps, xlab = "# Daily Steps", 
          main="Frequency of # of Daily Steps")
```

 - New mean and median are higher than original.
 - New histogram is more bell curved.
 - Both of these results are due to replacing the NAs with the average steps per interval.  
 
## Are there differences in activity patterns between weekdays and weekends?
 - Use updated copy of the original activity dataset
 - Add weekday column that contains the day of the week for that row's date

```{r}
     newactivity$weekday <- weekdays(as.Date(as.character(newactivity$date)))
```

 - Add weekend column that contains "weekday"/"weekend" factor variable
```{r}
      newactivity$weekend <- ifelse((newactivity$weekday=="Saturday"|
            newactivity$weekday=="Sunday"),"weekend", "weekday")
      newactivity$weekend <- as.factor(newactivity$weekend)
```

 - Create new aggregate of interval steps on the weekend factor column and interval
```{r}
      newintervalsteps <- aggregate(list(steps=newactivity$steps), 
            by = list(weekend = newactivity$weekend, interval = newactivity$interval), 
            FUN = "mean", 
            na.rm=TRUE)
```

### Plot out interval steps by weekend and weekday
 - It seems like there are more steps in the morning on weekdays 
 - Step distribution was more even over the weekends,.

```{r}
     library(lattice)
     xyplot(newintervalsteps$steps ~ newintervalsteps$interval | newintervalsteps$weekend, 
            layout = c(1, 2), 
            type="l")
```